---
title: "TAE - Unit 2 - Recitation"
author: "GT"
date: "April 22, 2016"
output: html_document
---

### The Data

In this recitation we will apply some of the ideas from Moneyball to data from the
National Basketball Association (NBA).

```{r}
NBA = read.csv("NBA_train.csv")
str(NBA)
```
The data contains data from all teams in season since 1980, except for ones with
less than 82 games.

- *SeasonEnd* is the year the season ended
- *Team* is the name of the team
- *Playoff* is a binary variable for whether or not a team made it to the playoffs
- *W* is the number of regulaar season wins
- *PTS* are the points scored during the regular season
- *oppPTS* are the opponent points scored during the regular season.
- *FG* is the number of _successful_ field goals (2pt and 3pt)
- *FGA* is the number of field goal _attempts_.
- *X2P* is the number of successful 2pt.
- *X2PA* is the number of attempted 2pt.
- *X3P* is the number of successful 3pt.
- *X3PA* is the number of attempted 3pt.
- *FT* is the number of successful free throws.
- *FTA* is the number of attempted free throws.
- *ORB* are offensive rebounds
- *DRB* are defensive rebounds.
- *AST* stands for assists.
- *STL* stands for steals.
- *BLK* stands for blocks.
- *TOV* stands for turnover.

The variables 2P, 2PA, 3P and 3PA in the csv file are modified during the 
import of the data by R. R doesn't like it when a variable begins with a number.
If a variable begins with a number it will put an 'X' in front of it.

### Playoffs and Wins

The goal of a basketball team is similar to that of a baseball team, making the playoffs.

How many games does a team need to win in order to make the playoffs ?

Let's use the table command to figure this out for the NBA.

```{r}
table(NBA$W, NBA$Playoffs)
```

We see that after about 45 wins, teams almost always make it to the playoffs. So it seems like a good goal
would be to try to win about 42 games.

In basketball, games are won by scoring more points than the other team. Can we use the difference between
points scored and points allowed throughout the regular season in order to predict the number of games that
a team will win?

Create a variable that is the difference between points scored and points allowed.

```{r}
NBA$PTSdiff = NBA$PTS - NBA$oppPTS
plot(NBA$PTSdiff, NBA$W)
```

This look like there's a strong linear relationship between these two variables

```{r}
WinsReg = lm(W ~ PTSdiff, data=NBA)
summary(WinsReg)
```

The first thing that we notice is that we've got very significant variables over here and R-squared = 0.9423,
which is very high.

Let's write down the equation that we found.

`W = 41 + 0.0326 * PTSdiff`

So in order to win at least 42 games and have good chance of making it to the playoffs, we can calculate
that

```{r}
(42-41)/0.0326 #30.67
```

So we need to score at least 31 more points than we allow in order to win at least 42 games.

### Points scored

Let's now build an equation to predict points scored. The dependent variable would be PTS and the
independent variable would be some of the common basketball statistics present in the dataset: number of
two-point field goal attempts, the number of three-point field goal attempts, offensive rebounds, defensive
rebounds, assist, steals, blocks, turnovers, free throw attempts.

```{r}
PointsReg = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + DRB + TOV + STL + BLK, data=NBA)
summary(PointsReg)
```

We can see that some variables are very significant, others are less significant, like STL. And other seem
to be not significant at all, like DRB, TOV, BLK. The R-squared is pretty good, 0.8992, so it shows that
there are a linear relationship between points and all basketball statistics.

Let's using the residual to calculate the sum of squared errors.

```{r}
SSE = sum(PointsReg$residuals^2) # 28394314
SSE 
```

This is very big, but not directly interpetable quantity. But we can also calculate the root mean 
squared error, RMSE, that is much more interpretable.

```{r}
RMSE = sqrt(SSE/nrow(NBA)) # 184.4
RMSE
``` 

So on average, we make an error of about 184.4 points. That seems like a lot, but you have to compare it to
the average number of points in a season.

```{r}
mean(NBA$PTS)  # 8370
```

So being off by about 184.4 points is really not so bad. But there is room for improvememt. Not all the
variables were significant. Let's try to remove some of the insignificant variables one at a time.

The first variable that we would remove is probably, TOV. This is because the p value for turnovers, 0.6859,
is the highest of all of the p values. That's means that turnovers is the least statistically significant
variable in the model.

```{r}
PointsReg2 = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + DRB + STL + BLK, data=NBA)
summary(PointsReg2)
```

The R-squared is 0.8991, it's almost identical to the previous model. So it seems that wer're justifid in
removing turnovers.

Let's see if we can remove another one of the insignificant variables. The next one that we want to remove
based on p-value is  defensive rebound DRB.

```{r}
PointsReg3 = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + STL + BLK, data=NBA)
summary(PointsReg3)
```

The R-squared it's the same, 0.8991. So it's justified again in removing defensive reobound.

Let's try one more time to remove a variable, in this case we can remove blocks, BLK.

```{r}
PointsReg4 = lm(PTS ~ X2PA + X3PA + FTA + AST + ORB + STL, data=NBA)
summary(PointsReg4)
```

The R-squared is still 0.8991 and the mode is a bit simpler, all the variables are significant.
Let's take a look at the sum of squared errors, SSE and RMSE.

```{r}
SSE_4 = sum(PointsReg4$residuals^2) # 28421465
SSE_4
RMSE_4 = sqrt(SSE_4/nrow(NBA)) # 184.5
RMSE_4
```

Essentially, we've kept the root mean squared error the same. So it seems like we've narrowed down a much
better model, because it's simpler and so more interpretable.

### Making Predictions

Now we'll try to make predictions for the 2012-2013 season.

Load the test set, because the training set only included data from 1980 until 2012. 

```{r}
NBA_test = read.csv("NBA_test.csv")
str(NBA_test)
```

Now we predict the points usingthe model made before.

```{r}
PointsPrediction = predict(PointsReg4, newdata = NBA_test)
```

Now that we have our prediction, how good is it ? We can compute the out-of-sample R-squared. 

First calculate the sum of squares that is the sum of the predicted amount minus the actual amount of points squared and summed. 
And we need to calcalate the total sum of squares that is the sum of the average number of points minus the
actual number of points.

```{r}
SSE = sum((PointsPrediction - NBA_test$PTS)^2)
SST = sum((mean(NBA$PTS) - NBA_test$PTS)^2)
R2 = 1 - (SSE/SST) # 0.8127
R2
```

We can also calculate the RMSE the same way as before.

```{r}
RMSE= sqrt(SSE/nrow(NBA_test)) # 196.37
RMSE
```

The value is a little bit higher than before, but it's not too bad.


