---
title: "TAE - Unit 2 - Lecture 2"
author: "GT"
date: "April 21, 2016"
output: html_document
---

## Unit 2 - Lecture 2 - Moneyball

Moneyball tells the story of the Oakland A's. The A's is a baseball team near
San Francisco, California. They were once a rich team, but the team was purchased in 
1995 by owners who enforced strict budget cuts.
In 2002, the A's lost three key players. By taking an analytics approach,
they were able to find underevalued players and form teams that were very efficient.
The A's selected players based on their statistics.

### Making it to the Playoffs

The goal of a baseball team is to make the playoffs. 

- How we can predict wheter or not a team make the playoffs by knowning how many
  games they won in the regular season.
- Use linear regression to predict how many games a team will win using the 
  difference between runs scored and runs allowed.
- Use linear regression to predict the number of runs a team will score using
  batting statistics and the number of runs a team will allow using fielding and
  pitching statistics
  
First question: How many games does a time need to win in the regular season to
make it to the playoff ?

In moneyball, Paul DePodesta judged that it would take 95 wins for the A's to make
it to the playoffs.

How does a team win games ?

They score more runs than their opponent does. The A's calculated that they need
to score 135 more runs than they allowed during the regular season to expect to win
95 games.

Let's see if we can verify this using linear regression in R.

```{r}
baseball = read.csv("baseball.csv")
str(baseball)
```

THis dataset includes an observation for every team and year pair from 1962 to 2012
for all seasons with 162 games.

Since we're confirming the claims made in Moneyball we want to build models using
Paul DePodesta data had in 2002, let's subsetting the data.

```{r}
moneyball = subset(baseball, Year < 2002)
str(moneyball)
```

In order to build a linear regression equation to predict wins using the difference
between runs scored and runs allowed, we'll create a new variable called, 'RD' for
Run Difference.

```{r}
moneyball$RD = moneyball$RS - moneyball$RA
str(moneyball)
plot(moneyball$RD, moneyball$W)
```

The scatter plot shows as that there's a very strong linear relationship between 
these two variables, which is a good sign for the linear regression equation.

Create now the linear regressino model, which we'll call WinsReg

```{r}
WinsReg = lm(W ~ RD, data=moneyball)
summary(WinsReg)
```

RD is very significant and the R-squared is 0.88. This is a strong model to predict
wins using difference between runs scored and runs allowed. 

From the model we can extract the equation: $W = 80.8814 + 0.1058*RD$.

We want Wins to be greater than or equal to 95: $W \geq 95$.

```{r}
RD = (95 - 80.8814) / 0.1058
RD
```

This is very close to the claim made in Moneyball that a team needs to score at
least 135 more runs than they allow to win at least 95 games.

### Predict runs

Now, we need to know how many runs a team will score, which we'll show can be predicted with batting
statistics, and how many runs a team will allow, wich can be predicted using fielding and pitching
statistics.

The Oakland A's were interested in answering the question: how does a team score more runs?
They discoverd two particular baseball statistics: on-base percentage, OBP, and slugging percentage, SLG.

- On-base percentage is the percentage of time a player gets on base, including walks.
- Slugging percentage measures how far a player gets around the bases on his turn and measures the power
  a hitter.
  
Most baseball teams and experts focused on batting average, which measures how often a hitter gets on base
by hitting the ball.

The Oakland A's claimed that on-base percentage was the most important statistic for predicting runs.

Let's see if we can use linear regression in R, to verify wich baseball statistic are important for
predicting runs scored.

```{r}
RunsReg = lm(RS ~ OBP + SLG + BA, data=moneyball)
summary(RunsReg)
```

We can see that all our independent variable are significant and the R-squared is 0.93. But if we look closer
at the coefficients, we can see that the coefficient for batting average is negative.

This implies that, a team with a lower batting average will score more runs, which is a little counterintuitive

This seems a case of multicollinearity. These three hitting statistics are highly correlated.

Let's try removing batting average to see what happens to our model.

```{r}
RunsReg = lm(RS ~ OBP + SLG, data=moneyball)
summary(RunsReg)
```

We can see that our independent variables are still very significant, the coefficients are both positive and
the R-squared is still about 0.93. 

This model is simpler, with only two independent variables and has about the same R-squared.

If we compare the two coefficient, we sse that on-base percentage has a larger coefficient than slugging
percentage. This tells us that on-base percentage is probabily worth more than slugging percentage.

So the claims made in Moneyball are verified.

We can create a very similar model to predict runs allowed. This model uses pitching statistics: 
opponents on-base percentage, OOBP, and opponents slugging percentage, OSLG.

```{r}
RunsAllReg = lm(RA ~ OOBP + OSLG, data=moneyball)
summary(RunsAllReg)
```

The key message here is that simple models, using only a couple independent variables, can be 
constructed to answer some of the most important questions in baseball.


### Using the Models to Make Predictions

We would like to predict before the season starts how many games the 2002 Oakland A's will win.

To do this, we first have to predict how many runs the team will score and how many runs they will allow.
These models use team statistics. However, when we are predicting for the 2002 Oakland A's before the season
has occurred, the team is probabily different than it was the year before.
 
So we don't know the team statistic. But we can estimate these statistics using past player performance
assuming that past performance correlates with future performance and that will be few injuries during
the season.

Let's start by making a prediction for runs scored.

At the beginning of the 2002 season, the Oakland A's had 24 batters on their roster. We can estimate that
team on-base percentage will be about 0.339 and team slugging percentage will be about 0.430.

Using the previous model, we predict that the 2002 Oakland A's will score about 805 runs.

Similarly, we can make a prediction for runs allowed. At the beginning of the 2002 season, the Oakland A's 
had 17 pitchers on their roster. Using the 2001 regular season statistics for these players, we can
estimate that team opponent on-base percentage will be about 0.307 and team opponent slugging percentage
will be about 0.373. 

Using the previous model, we predict that he 2002 Oakland A's will allow 622 runs.

We can plug the difference 805-622 in the previous model and predict that the A's will win 100 games in 2002.

These predictions are similar to Paul DePodesta's predictions. And it turns out that the 2002 Oakland A's
actually scored 800 and allowed 653 runs. And they won 103 games.

These predictions show us that by using publicly available data and simple analytics, we can predict
very close to what actually happened before the season even started.

 
