---
title: "TAE - Unit 2 - Assignment 2"
author: "GT"
date: "24 avril 2016"
output: html_document
---

## READING TEST SCORES

Each row in the datasets pisa2009train.csv and pisa2009test.csv represents one student taking the exam. 
The datasets have the following variables:

- *grade*: The grade in school of the student (most 15-year-olds in America are in 10th grade)
- *male*: Whether the student is male (1/0)
- *raceeth*: The race/ethnicity composite of the student
- *preschool*: Whether the student attended preschool (1/0)
- *expectBachelors*: Whether the student expects to obtain a bachelor's degree (1/0)
- *motherHS*: Whether the student's mother completed high school (1/0)
- *motherBachelors*: Whether the student's mother obtained a bachelor's degree (1/0)
- *motherWork*: Whether the student's mother has part-time or full-time work (1/0)
- *fatherHS*: Whether the student's father completed high school (1/0)
- *fatherBachelors*: Whether the student's father obtained a bachelor's degree (1/0)
- *fatherWork*: Whether the student's father has part-time or full-time work (1/0)
- *selfBornUS*: Whether the student was born in the United States of America (1/0)
- *motherBornUS*: Whether the student's mother was born in the United States of America (1/0)
- *fatherBornUS*: Whether the student's father was born in the United States of America (1/0)
- *englishAtHome*: Whether the student speaks English at home (1/0)
- *computerForSchoolwork*: Whether the student has access to a computer for schoolwork (1/0)
- *read30MinsADay*: Whether the student reads for pleasure for 30 minutes/day (1/0)
- *minutesPerWeekEnglish*: The number of minutes per week the student spend in English class
- *studentsInEnglish*: The number of students in this student's English class at school
- *schoolHasLibrary*: Whether this student's school has a library (1/0)
- *publicSchool*: Whether this student attends a public school (1/0)
- *urban*: Whether this student's school is in an urban area (1/0)
- *schoolSize*: The number of students in this student's school
- *readingScore*: The student's reading score, on a 1000-point scale

### Problem 1

```{r}
pisaTrain = read.csv("pisa2009train.csv")
str(pisaTrain)
summary(pisaTrain)
pisaTest = read.csv("pisa2009test.csv")
str(pisaTest)
tapply(pisaTrain$readingScore, pisaTrain$male, mean)
## remove all observation containig missing data
pisaTrain = na.omit(pisaTrain)
pisaTest = na.omit(pisaTest)
str(pisaTrain)
str(pisaTest)
```

### Problem 3

```{r}
# By default R selects the first level alphabetically as the reference level of the factor variable.
# Set the reference level of the factor by typing
pisaTrain$raceeth = relevel(pisaTrain$raceeth, "White")
pisaTest$raceeth = relevel(pisaTest$raceeth, "White")
lmScore = lm(readingScore ~ ., data = pisaTrain)
summary(lmScore) # R2 = 0.3251
SSE = sum((lmScore$residuals)^2)
SSE
RMSE = sqrt(SSE/nrow(pisaTrain)) # 73.36555
RMSE
(143.76+29.54*11) - (143.76+29.54*9) # difference between two students grade 11 and grade 9
```

### Problem 4

```{r}
predTest = predict(lmScore, newdata=pisaTest)
summary(predTest)
max(predTest)-min(predTest)
SSE = sum((predTest - pisaTest$readingScore)^2)
SSE
RMSE = sqrt(SSE/nrow(pisaTest)) # 73.36555
RMSE
SST = sum((mean(pisaTrain$readingScore) - pisaTest$readingScore)^2)
SST
R2 = 1 - (SSE/SST) # 0.2615
R2
# baseline
mean(pisaTrain$readingScore)
```



