---
title: "TAE - Unit 2 - Assignment 3"
author: "GT"
date: "April 25, 2016"
output: html_document
---

## DETECTING FLU EPIDEMICS VIA SEARCH ENGINE QUERY DATA

We would like to estimate influeny-like illness (ILI) activity using Google web 
search logs. 

The CDC publishes on its website the official regional and state-level percentage
of patients visits to healthcare providers for ILI purposes on a weekly basis.

Google trends allows public retrieval of weekly counts for every query searched by
users around the world.

The csv file FluTrain.csv aggregates this data from January 1, 2004 until December
31, 2011 as follows:

- *Week* - The range of dates represented by this observation, in year/month/day 
  format.
  
- *ILI* - This column lists the percentage of ILI-related physician visits for the
  corresponding week.

- *Queries* - This column lists the fraction of queries that are ILI-related for the
  corresponding week, adjusted to be between 0 and 1 (higher values correspond to
  more ILI-related search queries).
  
### Problem 1
  
```{r}
FluTrain = read.csv("FluTrain.csv")
str(FluTrain)
summary(FluTrain)
FluTrain$Week[which.max(FluTrain$ILI)]
FluTrain$Week[which.max(FluTrain$Queries)]
hist(FluTrain$ILI,breaks = 100)
plot(FluTrain$Queries,log(FluTrain$ILI))
```

### Problem 2

```{r}
FluTrend1 = lm(log(ILI) ~ Queries, data=FluTrain)
summary(FluTrend1)
cor(log(FluTrain$ILI), FluTrain$Queries)
0.842^2
log(1/0.842)
exp(-0.5*0.842)
```

### Problem 3

```{r}
FluTest = read.csv("FluTest.csv")
str(FluTest)
PredTest1 = exp(predict(FluTrend1, newdata=FluTest))
PredTest1[which(FluTest$Week == "2012-03-11 - 2012-03-17")] # 11
(FluTest$ILI[11]-PredTest1[11])/FluTest$ILI[11]
SSE = sum((FluTest$ILI - PredTest1)^2)
SSE
RMSE = sqrt(SSE/nrow(FluTest))
RMSE
```


