---
title: "TAE - Unit2 - Assignment 1"
author: "GT"
date: "23 avril 2016"
output: html_document
---

## CLIMATE CHANGE

The file climate_change.csv contains climate data from May 1983 to December 2008. The available variables include:

- *Year*: the observation year.
- *Month*: the observation month.
- *Temp*: the difference in degrees Celsius between the average global temperature in that period and a
  reference value.
- *CO2, N2O, CH4, CFC.11, CFC.12*: atmospheric concentrations of carbon dioxide (CO2), nitrous oxide (N2O),
  methane  (CH4), trichlorofluoromethane (CCl3F; commonly referred to as CFC-11) and dichlorodifluoromethane
  (CCl2F2; commonly referred to as CFC-12), respectively. 
    + CO2, N2O and CH4 are expressed in ppmv (parts per million by volume  -- i.e., 397 ppmv of CO2 means t
      hat CO2 constitutes 397 millionths of the total volume of the atmosphere)
    + CFC.11 and CFC.12 are expressed in ppbv (parts per billion by volume). 
- *Aerosols*: the mean stratospheric aerosol optical depth at 550 nm. This variable is linked to volcanoes, as
  volcanic eruptions result in new particles being added to the atmosphere, which affect how much of the sun's
  energy is reflected back into space.
- *TSI*: the total solar irradiance (TSI) in W/m2 (the rate at which the sun's energy is deposited per unit
  area). Due to sunspots and other solar phenomena, the amount of energy that is given off by the sun varies
  substantially with time. 
- *MEI*: multivariate El Nino Southern Oscillation index (MEI), a measure of the strength of the El Nino/La
  Nina-Southern Oscillation (a weather effect in the Pacific Ocean that affects global temperatures). 
  
### Problem 1

Read the dataset climate_change.csv into R. Then split the data into a *training* set, consisting of all obs
up to and including 2006 and a *testing* set, consisting of the remaining years. Next build a linear
regression model to predict the dependent variable Temp, using MEI, CO2, CH4, N20, CFC.11, CFC.12, TSI, Aerosols

```{r}
CC = read.csv("climate_change.csv")
str(CC)
CC_training = subset(CC, Year <= 2006)
str(CC_training)
CC_testing = subset(CC, Year > 2006)
str(CC_testing)
CCReg = lm(Temp ~ MEI + CO2 + CH4 + N2O + CFC.11 + CFC.12 + TSI + Aerosols, data = CC_training)
summary(CCReg) # Multiple R-squared:  0.7509
# The following variables are significative: MEI, CO2, CFC.11, CFC.12, TSI, Aerosols
```

### Problem 2

Multicollinarity between N20 and CFC.11.

```{r}
cor(CC_training)
```

### Problem 3

Build a model with only MEI, TSI, Aerosols and N2O

```{r}
CCReg2 = lm(Temp ~ MEI + N2O + TSI + Aerosols, data = CC_training)
summary(CCReg2) # Multiple R-squared: 0.7261
```

### Problem 4

```{r}
CCReg3 = step(CCReg)
summary(CCReg3) # Multiple R-squared: 0.7508, the step function eliminates CH4
```

### Problem 5

```{r}
CCPredict = predict(CCReg3, newdata = CC_testing)
CCPredict
SSE = sum((CC_testing$Temp - CCPredict)^2)
SSE
SST = sum((CC_testing$Temp - mean(CC_training$Temp))^2)
SST
R2 = 1-SSE/SST
R2
```

