---
title: "TAE - Unit 7 - Lecture 1"
author: "GT"
date: "May 24, 2016"
output: html_document
---

## Visualizing the Word

In this lecture, we'll discuss the idea of using visualization to better understand
data and to provide insights on the problem we're addressing.

*Why visualization?*

People often say that a picture is like a thousand words.

In the same spirit, John Tukey, a major statistician at Princeton, 
wrote that 

>  "the picture-examining eye is the best finder we have of the 
   wholly unanticipated."
  
Visualizing data allows us to discern relationships, structures, distributions,
outliers, patterns, behaviors, dependencies, and outcomes.

Visualization is further useful for initial data exploration, for interpreting 
models, and for communicating results effectively.

We'll see how visualizations can be used to better understand data, communicate
information more effectively, show the results of analytical models.

### The World Health Organization (WHO)

The World Health Organization, WHO for short, is the authority for health within 
the United Nations system.

It is responsible for providing leadership on global health matters, shaping the
health research agenda, setting norms and standards, articulating evidence-based
policy options, providing technical support to countries, and monitoring and 
assessing health trends.

WHO communicates information about global health in order to inform citizens,
donors, policymakers, and organizations around the world.

Their primary publication is World Health Report.

Each issue focuses on a specific aspect of global health, and includes statistics
and experts' assessments. WHO also maintains an open, online repository
of global health data. WHO provides some data visualizations, which helps them
communicate more effectively with the public.

### What is Data Visualization ?

Data visualization is defined as a mapping of data properties to visual properties.

Data properties are usually numerical or categorical, like the mean of a variable,
the maximum value of a variable, or the number of observations with a certain
property.

Visual properties can be (x,y) coordinates to plot points on a graph, colors to
assign labels, sizes, shapes, heights, etc.

Both types of properties are used to better understand the data, but in 
different ways.

To motivate the need for data visualization, let's look at a famous example 
called Anscombe's Quartet.

Each of these tables corresponds to a different data set. We have four data sets,
each with two variables, x and y.

```
|  X1  |   Y1  |  |  X2  |  Y2  |  |  X3  |   Y3  |   |  X4  |   Y4  |
|------+-------|  |------+------|  |------+-------|   |------+-------|
| 10.0 |  8.04 |  | 10.0 | 9.14 |  | 10.0 |  7.46 |   |  8.0 |  6.58 | 
|  8.0 |  6.95 |  |  8.0 | 8.14 |  |  8.0 |  6.77 |   |  8.0 |  5.76 |
| 13.0 |  7.58 |  | 13.0 | 8.74 |  | 13.0 | 12.74 |   |  8.0 |  7.71 |
|  9.0 |  8.81 |  |  9.0 | 8.77 |  |  9.0 |  7.11 |   |  8.0 |  8.84 |
| 11.0 |  8.33 |  | 11.0 | 9.26 |  | 11.0 |  7.81 |   |  8.0 |  8.47 |
| 14.0 |  9.96 |  | 14.0 | 8.10 |  | 14.0 |  8.84 |   |  8.0 |  7.04 |
|  6.0 |  7.24 |  |  6.0 | 6.13 |  |  6.0 |  6.08 |   |  8.0 |  5.25 |
|  4.0 |  4.62 |  |  4.0 | 3.10 |  |  4.0 |  5.39 |   |  8.0 | 12.50 |  
| 12.0 | 10.84 |  | 12.0 | 9.13 |  | 12.0 |  8.15 |   | 19.0 |  5.56 |
|  7.0 |  4.82 |  |  7.0 | 7.26 |  |  7.0 |  6.42 |   |  8.0 |  7.91 |  
|  5.0 |  5.68 |  |  5.0 | 4.74 |  |  5.0 |  5.73 |   |  8.0 |  6.89 |

Mean of X: 9.0          Mean of Y: 7.50         Correl(X,Y): 0.816
Variance of X: 11.0     Variance of Y: 4.12     Regression eq.: Y=3.00+0.50X
```

Just looking at the tables of data, it's hard to notice anything special about it.
It turns out that the mean and variance of the x variable is the same for all four
data sets, the mean and variance of the y variable is the same for all four data
sets, and the correlation between x and y, as well as the regression equation to
predict y from x, is the exact same for all four data sets.

So just by looking at data properties, we might conclude that these data sets 
are very similar.

But if we plot the four data sets, they're very different.

```
y1^                                     y2^
  |                 *    *                |
  |                                       |             
  |             *    *                    |           *  *
  |         *    *                        |         *       *
  |    * *                                |       *          *
  |   *   *                               |      *
  |  *                                    |     *
  +---------------------------->          +---------------------------->
                               x1                                      x2
y3^                                     y4^
  |                  *                    |                         *
  |                                       |             
  |                +  *                   |      *    
  |           +*                          |      *
  |      +*                               |      *
  |   *                                   |      *
  |                                       |     
  +---------------------------->          +---------------------------->
                               x3                                      x4
```


These plots show the four data sets, with the x variable on the x-axis,
and the y variable on the y-axis.

Visually, these data sets look very different. But without visualizing them, 
we might not have noticed this.

This is one example of why visualizing data can be very important.

We'll use the **ggplot2** package in R to create data visualizations.

This package was created by Hadley Wickham, who described ggplot as 

> "a plotting system for R based on the grammar of graphics, which
tries to take the good parts of base and lattice graphics and none 
of the bad parts. It takes care of many of the fiddly details that
make plotting a hassle (like drawing legends) as well as providing a 
powerful model of graphics that makes it easy to produce complex 
multi-layered graphics."

So what do we gain from using ggplot over just making plots using the basic R
functions, or what's referred to as base R?

Well, in base R, each mapping of data properties to visual properties is its 
own special case. 

When we create a scatter plot, or a box plot, or a histogram, we have to use a
completely different function. Additionally, the graphics are composed of simple
elements, like points or lines.

It's challenging to create any sophisticated visualizations. It's also difficult 
to add elements to existing plots.

But in ggplot, the mapping of data properties to visual properties is done by
just adding layers to the plot. This makes it much easier to create sophisticated
plots and to add to existing plots.

So what is the grammar of graphics that ggplot is based on?

All ggplot graphics consist of three elements.

1. **Data**, in a data frame.
2. **Aesthetic mapping**, which describes how variables in the data frame
   are mapped to graphical attributes. 
   *Color, shape, scale, x-y axes, subsets,...*
3. **Geometric objects** determine how the data values are rendered graphically.
   *Points, lines, bars, boxes, etc.*

### Basic Scatterplots Using ggplot

Let's start by reading in our data. 

```{r}
WHO = read.csv("WHO.csv")
str(WHO)
```

We can see that we have 194 observations, or countries, and 13 different 
variables.

In week one, the very first plot we made in R was a scatterplot of fertility rate
versus gross national income.

Let's make this plot again, just like we did in week one.

```{r}
plot(WHO$GNI, WHO$FertilityRate)
```

This plot shows us that a higher fertility rate is correlated with a lower income.

Now, let's redo this scatterplot, but this time using ggplot. We'll see how ggplot
can be used to make more visually appealing and complex scatterplots.

First, we need to install and load the ggplot2 package.

```{r}
library(ggplot2)
```

Now, remember we need at least three things to create a plot using ggplot, 
data, an aesthetic mapping of variables in the data frame to visual output,
and a geometric object.

So first, let's create the ggplot object with the data and the aesthetic mapping.

```{r}
scatterplot = ggplot(WHO, aes(x = GNI, y = FertilityRate))
```

Now, we need to tell ggplot what geometric objects to put in the plot. We could use
bars, lines, points, or something else.

This is a big difference between ggplot and regular plotting in R. You can build
different types of graphs by using the same ggplot object.

So first, let's just create a straightforward scatterplot.

```{r}
scatterplot + geom_point()
```

You should see a new plot in the Graphics window that looks similar to our 
original plot, but there are already a few nice improvements.

One is that we don't have the data set name with a dollar sign in front of the 
label on each axis, just the variable name. Another is that we have these nice
grid lines in the background and solid points that pop out from the background.

We could have made a line graph just as easily by changing point to line.

```{r}
scatterplot + geom_line()
```

Now, you can see a line graph in the Graphics window. However, a line doesn't 
really make sense for this particular plot. 

In addition to specifying that the geometry we want is points, we can add other
options, like the color, shape, and size of the points.

Let's redo our plot with blue triangles instead of circles. 

```{r}
scatterplot + geom_point(color="blue", size=3, shape=17)
scatterplot + geom_point(color="darkred", size=3, shape=8)
```

There are many different colors and shapes that you can specify.

Now, let's add a title to the plot.

```{r}
scatterplot + geom_point(color="darkred", size=3, shape=8) + 
  ggtitle("Fertility Rate vs. Gross National Income")
```

You should now see that it has a nice title at the top. 

Now, let's save our plot to a file.

We can do this by first saving our plot to a variable.

```{r}
fertilityGNIplot = scatterplot + geom_point(color="darkred", size=3, shape=8) + 
  ggtitle("Fertility Rate vs. Gross National Income")
```

This will save our scatterplot to the variable, fertilityGNIplot.

Now, let's create a file we want to save our plot to. We can do that with the pdf
function.

```{r}
pdf("MyPlot.pdf")
print(fertilityGNIplot)
dev.off() # to close the file
```

Now, if you look at the folder where WHO.csv is, you should see another file
called MyPlot.pdf, containing the plot we made.

### Advanced Scatterplots Using ggplot


