---
title: "TEA - Unit 7 - Lecture 2"
author: "GT"
date: "25 mai 2016"
output: html_document
---

##  The Analytical Policeman

In this lecture, we will discuss how visualization can offer insights in the area of policing
in urban environments.

The explosion of computerized data affects all parts of society, including policing.

In the past, human judgment and experience was the only tool in identifying patterns in criminal 
behavior. 

Police forces around the United States and the world are augmenting human judgment with
analytics - sometimes described as **predictive analysis**.

These days the LAPD is on an offensive to prevent crime. Its latest weapon is a computer program
that can actually predict where crimes will happen.

In the Foothill Division north of downtown Los Angeles, police are patrolling the largely
working-class neighborhoods with specially marked maps. The small red squares are hot spots,
where computers project property crimes are most likely.

It's called predictive policing.

These crime prediction boxes come from the same kind of mathematical calculation used
to predict earthquakes and aftershocks. By analyzing the times, dates, and places of recent crimes
computers project hot spots for burglaries, break-ins, and car thefts.

LA's Police Chief Charlie Beck says increasing police patrols inside those boxes
denies criminals opportunity.The real measure of this is not how many people you catch,
it's how much crime you prevent.

The LAPD began testing the predictive policing model here in the Foothill Division in November,
and the early results are encouraging. Burglaries are down 33%, and violent crime is also down 21%.
T
hat success will allow Beck to expand the program to other parts of the city and leverage 
limited resources.

The Los Angeles Police Chief Charlie Beck writes, 

> "I'm not going to get more money. I'm not going to get more cops. I have to be better at using 
  what I have, and that's what predictive policing is about. If this old street cop can change the way
  that he thinks about this stuff, then I know that my officers can do the same."

Let me comment on the role of analytics.

The analytical tools you have learned in this class can be used to make these predictive policing 
models possible.

However, *communicating* the results of these models is critical. A *linear* regression output table
will not be of use to a *police* person on patrol.

Visualization bridges the gap between the *analytics* and the *end user*.

### Visualizing Crime Over Time

In almost every application, before we even consider a predictive model, we should 
try to understand the historical data.

Many cities in the United States and around the world, provide logs of reported crimes, 
usually the time, location, and nature of the event.

In this lecture, we'll use data from the city of Chicago, in the United States, 
about motor vehicle thefts.

Given this data on crimes, suppose we wanted to communicate crime patterns
over the course of an average week.

We could display daily crime averages using a line graph, like the one shown here, 
but this doesn't seem too useful.

We can see that crime tends to be higher on Saturday, but when on Saturday, and where?

We could replace our x-axis with the hour of the day and have a different line for
every day of the week to understand when crime occurs in more detail.

But this would be a jumbled mess with seven lines, and probably very hard to read.

We could instead use no visualization at all, and instead present information in a 
table.

For each hour and day, we have the total number of crimes that occurred. This is a 
valid representation of the data, but large tables of numbers can be hard to read and
understand.

So how can we make the table more interesting and usable?

A great way to visualize information in a two-dimensional table is with a **heatmap**.

**Heatmaps** visualize data using three attributes. Two of the attributes are on the 
*x and y-axes*, typically displayed horizontally and vertically.

The *third attribute* is represented by **shades of color**.

In this example, lower values in the third attribute correspond to colors closer to blue,
and higher values in the third attribute correspond to colors closer to red.

For example, the x-axis could be hours of the day, the y-axis could be days of the week,
and the colors could correspond to the amount of crime.

In a heat map, we can pick different color schemes based on the type of data to convey 
different messages.

In crime, a yellow to red color scheme might be appropriate because it can highlight
some of the more dangerous areas in red.

Your eye is naturally drawn to the red areas of the plot.

In other applications, both high and low values are meaningful, so having a more varied 
color scheme might be useful.

And in other applications, you might only want to see cells with high values,
so you could use a gray scale to make the cells with low values white.

The x and y-axes in a heat map don't need to be continuous. 

In our example, we have a categorical or factor variable, the day of the week.

And we can even combine a heat map with a geographical map, which we'll discuss 
later in this lecture.

This type of heat map is frequently used in predictive policing to show crime hot 
spots in a city.

In this lecture, we'll use Chicago motor vehicle theft data to explore patterns of
crime, both over days of the week, and over hours of the day.

We're interested in analyzing the total number of car thefts that occur in any 
particular hour of a day of the week over our whole data set.

### A Line Plot

Let's start by reading in our data. 

```{r}
mvt = read.csv("mvt.csv", stringsAsFactors = FALSE) # since we have a text field
str(mvt)
```

We have over 190,000 observations of three different variables:

- the date of the crime
- the location of the crime, in terms of *latitude* and *longitude*.

We want to first convert the Date variable to a format that R will recognize so that we
can extract the day of the week and the hour of the day.

We can do this using the strptime function. So we want to replace our variable, Date,
with the output of the strptime function, which takes as a first argument our variable, 
Date, and then as a second argument the format that the date is in.

```{r}
mvt$Date = strptime(mvt$Date, format="%m/%d/%y %H:%M")
```

In this format, we can extract the hour and the day of the week from the Date variable,
and we can add these as new variables to our data frame. We can do this by first defining 
our new variable, Then, we just take the hour variable out of Date variable.

```{r}
Sys.setlocale("LC_ALL", "C")
mvt$Weekday = weekdays(mvt$Date)
mvt$Hour = mvt$Date$hour
str(mvt)
```

Now, we have two more variables:

- Weekday, which gives the day of the week
- Hour, which gives the hour of the day.

Now, we're ready to make some line plots. Let's start by creating the line plot
 with just one line and a value for every day of the week.

We want to plot as that value the total number of crimes on each day of the week.
We can get this information by creating a table of the Weekday variable.

```{r}
table(mvt$Weekday)
```

This gives the total amount of crime on each day of the week.

Let's save this table as a data frame so that we can pass it to ggplot as our data.

```{r}
WeekdayCounts = as.data.frame(table(mvt$Weekday))
str(WeekdayCounts)
```

We can see that our data frame has seven observations, one for each day of the week, 
and two different variables. The first variable, called Var1, gives the name of the day
of the week, and the second variable, called Freq, for frequency, gives the total 
amount of crime on that day of the week.

Now, we're ready to make our plot. First, we need to load the ggplot2 package.

```{r}
library(ggplot2)
```

Now, we'll create our plot using the ggplot function.

```{r}
ggplot(WeekdayCounts, aes(x = Var1, y = Freq)) + geom_line(aes(group=1)) # this groups all of data into one line
```

We can see that this is very close to the plot we want. We have the total number of crime
plotted by day of the week, but our days of the week are a little bit out of order.

We have Friday first, then Monday, then Saturday, then Sunday, etc. What ggplot did was
it put the days of the week in alphabetical order.

But we actually want the days of the week in chronological order to make this plot a bit 
easier to read. We can do this by making the Var1 variable an ordered factor variable.
This signals to ggplot that the ordering is meaningful.

We can do this by using the factor function where the first argument is our variable,
WeekdayCounts$Var1, the second argument is ordered = TRUE. And the third argument, which 
is levels,should be equal to a vector of the days

```{r}
WeekdayCounts$Var1 = factor(WeekdayCounts$Var1, ordered = TRUE, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
ggplot(WeekdayCounts, aes(x = Var1, y = Freq)) + geom_line(aes(group=1))
```

Now, this is the plot we want. We have the total crime by day of the week with the days 
of the week in chronological order.

The last thing we'll want to do to our plot is just change the x- and y-axis labels,
since they're not very helpful as they are now. 

```{r}
ggplot(WeekdayCounts, aes(x = Var1, y = Freq)) + geom_line(aes(group=1)) + xlab("Day of the Week") + ylab("Total Motor Vehicle Theft")
```

### A Heatmap

We'll add the hour of the day to our line plot, and then create an alternative visualization 
using a heat map. 

We can do this by creating a line for each day of the week and making the x-axis the hour of the day.

We first need to create a counts table for the weekday, and hour.

```{r}
table(mvt$Weekday, mvt$Hour)
```

This table gives, for each day of the week and each hour, the total number of motor vehicle thefts 
that occurred.

Let's save this table to a data frame so that we can use it in our visualizations.
We'll call it DayHourCounts and use the as.data.frame function, run on our table, where the 
first variable is the Weekday and the second variable is the Hour.

```{r}
DayHourCounts = as.data.frame(table(mvt$Weekday, mvt$Hour))
str(DayHourCounts)
```

We can see that we have 168 observations, one for each day of the week and hour pair, and three 
different variables:

- *Var1* gives the day of the week.
- *Var2* gives the hour of the day.
- *Freq* for frequency gives the total crime count.

Let's convert the second variable, Var2, to actual numbers and call it Hour, since this is the hour of
the day, and it makes sense that it's numerical.

```{r}
DayHourCounts$Hour = as.numeric(as.character(DayHourCounts$Var2))
str(DayHourCounts)
```

**This is how we convert a factor variable to a numeric variable.**

Now we're ready to create our plot. We just need to change the group to Var1, which is the day 
of the week. 

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Freq)) + geom_line(aes(group = Var1))
```

You should see a new plot show up in the graphics window. It has seven lines, one for each day 
of the week. While this is interesting, we can't tell which line is which day, so let's change
the colors of the lines to correspond to the days of the week.

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Freq)) + geom_line(aes(group = Var1, color = Var1), size = 2)
```

Now in our plot, each line is colored corresponding to the day of the week. This helps us see that on 
Saturday and Sunday, for example, the green and the teal lines, there's less motor vehicle thefts in 
the morning.

While we can get some information from this plot, it's still quite hard to interpret. Seven lines 
is a lot. Let's instead visualize the same information with a heat map.

To make a heat map, we'll use our data in our data frame DayHourCounts. First, though, we need to 
fix the order of the days so that they'll show up in chronological order instead of in alphabetical
order.

So for DayHourCounts$Var1, which is the day of the week, we're going to use the factor function.

```{r}
DayHourCounts$Var1 = factor(DayHourCounts$Var1, ordered = TRUE, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Now let's make our heat map.

`geom_tile()`  is the function we use to make a heat map. And then in the aesthetic for our tiles,
we want the fill to be equal to Freq. This will define the colors of the rectangles in our heat map
to correspond to the total crime.

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Var1)) + geom_tile(aes(fill = Freq))
```

You should see a heat map pop up in your graphics window.

So how do we read this? For each hour and each day of the week, we have a rectangle in our heat map.
The color of that rectangle indicates the frequency,or the number of crimes that occur in that hour
and on that day. Our legend tells us that lighter colors correspond to more crime. So we can see that 
a lot of crime happens around midnight, particularly on the weekends.

We can change the label on the legend, and get rid of the y label to make our plot a little nicer.

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Var1)) + geom_tile(aes(fill = Freq)) + scale_fill_gradient(name="Total MV Thefts") + theme(axis.title.y = element_blank())
```

And now on our heat map, the legend is titled "Total MV Thefts" and the y-axis label is gone.

We can also change the color scheme. We'll make lower values correspond to white colors and
higher values correspond to red colors.

```{r}
ggplot(DayHourCounts, aes(x = Hour, y = Var1)) + geom_tile(aes(fill = Freq)) + scale_fill_gradient(name="Total MV Thefts", low="white", high="red") + theme(axis.title.y = element_blank())
```

This is a common color scheme in policing. It shows the hot spots, or the places with more 
crime, in red. So now the most crime is shown by the red spots and the least crime is shown
by the lighter areas.

It looks like Friday night is a pretty common time for motor vehicle thefts.

We saw something that we didn't really see in the heat map before.

It's often useful to change the color scheme depending on whether you want high values or 
low values to pop out, and the feeling you want the plot to portray.

### A Geographical Hot Spot Map

In this video, we'll plot crime on a map of Chicago.
First, we need to install and load two new packages, the maps
package and the ggmap package.
So start by installing the package maps.
So type install.packages("maps").
When the CRAN mirror window pops up,
go ahead and pick a location near you.
When the package is done installing and you're back
at the blinking cursor, also type install.packages("ggmap").
When that package is also done installing,
load both packages using the library command.
So type library(maps), and then library(ggmap).
Now, let's load a map of Chicago into R.
We can easily do this by using the get_map function.
So we'll call it chicago = get_map(location = "chicago",
zoom = 11).
Let's take a look at the map by using the ggmap function.
Now, in your R graphics window, you
should see a geographical map of the city of Chicago.
Now let's plot the first 100 motor vehicle
thefts in our data set on this map.
To do this, we start by typing ggmap(chicago).
This is instead of using ggplot like we've
been using in the previous videos.
Then we want to add geom_point, and here, we'll
define our data set to be equal to motor vehicle thefts, where
we'll take the first through 100th observations,
and in our aesthetic, we'll define our x-axis
to be the longitude of the points and our y-axis
to be the latitude of the points.
Now, in your R graphics window, you
should see the map of Chicago with black points marking where
the first 100 motor vehicle thefts were.
If we plotted all 190,000 motor vehicle thefts,
we would just see a big black box,
which wouldn't be helpful at all.
We're more interested in whether or not
an area has a high amount of crime,
so let's round our latitude and longitude
to two digits of accuracy and create a crime counts data
frame for each area.
We'll call it LatLonCounts, and use the as.data.frame function
run on the table that compares the latitude and longitude
rounded to two digits of accuracy.
So our first argument to table is round(mvt$Longitude, 2).
And our second argument is round(mvt$Latitude, 2).
This gives us the total crimes at every point on a grid.
Let's take a look at our data frame using the str function.
We have 1,638 observations and three variables.
The first two variables, Var1 and Var2,
are the latitude and longitude coordinates,
and the third variable is the number of motor vehicle thefts
that occur in that area.
Let's convert our longitude and latitude variables to numbers
and call them Lat and Long.
So first, we'll define the variable in our LatLonCounts
data frame, called Long, and set that equal to as.numeric, run
on as.character.
Remember, this is how we convert a factor variable
to a numerical variable.
And we'll give the variable, LatLonCounts$Var1.
Now let's just repeat this for latitude.
So LatLonCounts$Lat = as.numeric(as.character(LatLonCounts$Var2)).
Now, let's plot these points on our map,
making the size and color of the points
depend on the total number of motor vehicle thefts.
So first, again we type ggmap(chicago) +
geom_point(LatLonCounts, aes(x = Long, y = Lat, color = Freq,
size = Freq)).
Now, in our R graphics window, our plot
should have a point for every area defined by our latitude
and longitude areas, and the points
have a size and color corresponding
to the number of crimes in that area.
So we can see that the lighter and larger points correspond
to more motor vehicle thefts.
This helps us see where in Chicago more crimes occur.
If we want to change the color scheme,
we can do that too by just hitting the up arrow in our R
console and then adding scale_color_gradient(low="yellow",
high="red").
If you hit Enter, you should see the same plot as before,
but this time, the areas with more crime are closer to red
and the areas with less crime are closer to yellow.
We can also use geom_tile to make something that looks more
like a traditional heat map.
To do this, we type ggmap(chicago), just
like before, but now we're going to use geom_tile, where
our data frame again is LatLonCounts.
And in our aesthetic, we have that the x-axis is Long,
the y-axis is Lat, and then we have alpha=Freq.
This will define how to scale the colors on the heat map
according to the crime counts.
Then close the parentheses and type a comma,
and then type fill="red", defining our color scheme.
Close the parentheses and hit Enter.
This map takes a minute to load.
While we're waiting, let's discuss
what we've done in this video.
We've created a geographical heat map, which in our case
shows a visualization of the data,
but it could also show the predictions of a model.
Now that our heat map is loaded, let's take a look.
In each area of Chicago, now that area
is colored in red by the amount of crime there.
This looks more like a map that people
use for predictive policing.
In the next video, we'll use data from the FBI
to make a heat map on a map of the United States.