---
title: "Unit 1 - Assignement2"
author: "GT"
date: "16 avril 2016"
output: html_document
---

```{r}
Sys.setlocale("LC_ALL", "C")
```

## STOCK DYNAMICS

Download and read the files in R, using the read.csv() function. Each data frame has two variables, described
as follows:

  - **Date**: thedate of the stock price (first of the month)
  - **StockPrice**: the average stock price of the company in the given month

```{r}
IBM = read.csv("IBMStock.csv")
GE = read.csv("GEStock.csv")
ProcterGamble = read.csv("ProcterGambleStock.csv")
CocaCola = read.csv("CocaColaStock.csv")
Boeing = read.csv("BoeingStock.csv")
```

### Problem 1 - Summary Statistics

Look at the structure of the datasets. The date variable is stored as a factor. We can convert this to a "Date"
object in R

```{r}
str(IBM)
IBM$Date = as.Date(IBM$Date, "%m/%d/%y")
GE$Date = as.Date(GE$Date, "%m/%d/%y")
CocaCola$Date = as.Date(CocaCola$Date, "%m/%d/%y")
ProcterGamble$Date = as.Date(ProcterGamble$Date, "%m/%d/%y")
Boeing$Date = as.Date(Boeing$Date, "%m/%d/%y")
str(IBM)
summary(IBM)
summary(GE)
summary(CocaCola)
summary(Boeing)
sd(ProcterGamble$StockPrice)
```

The five datasets have the same number of observations. How many observations are there in each data set ?
**480**

What is the earliest year in the datasets? **1970**

What is the latest year in the datasets? **2009**

What is the mean stock price of IBM over this time period? **144.38**

What is the minimum stock price of GE over this time period? **9.294**

What is the maximum stock price of Coca-Cola over this time period? **146.58**

What is the median stock price of Boeing over this time period? **44.88**

What it the standard deviation of the stock price of Procter&Gamble over this time period? **18.194**

### Problem 2 - Visualizing Stock Dynamics

Plot the stock prices to see if we can visualize trends in stock prices during this time period. 
Add the argument type = "l" to plot a line intead of points

```{r}
plot(CocaCola$Date, CocaCola$StockPrice, type="l")
```

Around what year did Coca-cola has its highest stock price in this time period ? **1973**

Around what year did Coca/Cola has its lowest stock price in this time period ? **1980**

Let's add the line for Procter & Gamble too. Add the line to a plot in R by using the lines function.
Add a color for the lines to distinguish them. Draw a vertical line at a certain date with the function
abline()

```{r}
plot(CocaCola$Date, CocaCola$StockPrice, type="l", col="red")
lines(ProcterGamble$Date, ProcterGamble$StockPrice, col="blue")
abline(v=as.Date(c("2000-03-01")))
abline(v=as.Date(c("1983-01-01")))
```

In March of 2000, the technology bubble burst, and a stock market crash occurred. According to this plot,
which company's stock dropped more? **Procter & Gamble**

Around 1983, the stock for one of these companies (CocaCola or Procter & Gamble) was goinig up, while the
other was going down. Which one was going up? **Coca-Cola**

In the time period shown in the plot, which stock generally has lower values? **Coca-Cola**

### Problem 3 - Visualizing Stock Dynamics 1995-2005

Let's take a look at how the stock prices changed from 1995-2005 for all five companies.

```{r}
plot(CocaCola$Date[301:432], CocaCola$StockPrice[301:432],type="l",col="red",ylim=c(0,210))
lines(ProcterGamble$Date[301:432], ProcterGamble$StockPrice[301:432],col="blue")
lines(IBM$Date[301:432], IBM$StockPrice[301:432], col="green")
lines(GE$Date[301:432], GE$StockPrice[301:432], col="brown")
lines(Boeing$Date[301:432], Boeing$StockPrice[301:432], col="black")
abline(v=as.Date(c("2000-03-01")))
abline(v=as.Date(c("1997-10-01")))
```

Which stock fell the most right after thge technology bubble burst in March 2000? **GE**

Which stock reaches the highest value in the time period 1995-2005? **IBM**

In October 1997 there was a global stock market crash that was caused by an economic crisis in Asia.
comparing September 1997 to November 1997, which companies saw a decreasing trend in their stock price? 
**Procter & Gamble and Boeing**

In the last two years of this time period (2004 and 2005) which stock seems to be performing the best, in 
terms of increasing stock price? **Boeing**

### Problem 4 - Monthly Trends

Let's see if stocks tend to be higher or lower during certain months. Use the tapply command to calculate
the mean stock price of IBM, sorted by months.

```{r}
mean(IBM$StockPrice)
tapply(IBM$StockPrice, months(IBM$Date), mean)
```

In which months has IBM historically had a higher stock price? **Jan, Feb, Mar, Apr, May**

```{r}
tapply(CocaCola$StockPrice, months(CocaCola$Date), mean)
tapply(GE$StockPrice, months(GE$Date), mean)
tapply(ProcterGamble$StockPrice, months(ProcterGamble$Date), mean)
tapply(Boeing$StockPrice, months(Boeing$Date), mean)
```

GE and Coca-Cola both have their highest average stock price in the same month. Which month is this? **April**

For the months of December and January, every company's averages stock is higher in one month and lower in the 
other. In which month are the stock prices lower? **December**
