---
title: "Unit 1  - Assignement3"
author: "GT"
date: "17 avril 2016"
output: html_document
---

## DEMOGRAPHICS AND EMPLOYMENT IN THE UNITED STATES

In the USA the governement measures unemployment using the Current Population Survey (CPS), which collecs
demographic and employment information from a wide range of Americans each month.

The obs in the dataset represent people surveyed in the September 2013 CPS who actually completed a survey.
The compact version of this database has the following variables:

  - **PeopleInHousehold**: The number of people in the interviewee's houshold
  - **Region**
  - **State**
  - **MetroAreaCode**: A code that identifies the metropolitan area in which the interviewee lives. The mapping is provided in the file MetroAreaCode.csv
  - **Age**
  - **Married**
  - **Sex**
  - **Education**
  - **Race**
  - **Hispanic**
  - **CountryOfBirthCode**: A code identifying the country of birth of the interviewee. The mapping from codes to names of countries is provided in the file CountryCodes.csv
  - **Citizenship**
  - **EmploymentStatus**
  - **Industry**

### Problem 1

Load the dataset from CPSData.csv into a data frame called CPS and view the dataset with the summary() and str()

```{r}
CPS = read.csv("CPSData.csv")
str(CPS)
summary(CPS)
sort(table(CPS$Industry))
sort(table(CPS$State))
table(CPS$Citizenship)
table(CPS$Hispanic, CPS$Race)
```

How many interviewees are in the dataset? **131302**

Among the interviewees with a value reported for the Industry variable, what is the most common industry of 
employment: **Educational and health services**

Which state has the fewest interviewees ? **New Mexico**

Which state has the largest number of interviewees ? **California**

What proportion of interviewees are citiziens of the United States? **`r (116639+7073)/131302 # 0.942`** 

For which races are there at least 250 interviewees in the CPS dataset of Hispanic ethnicity? 
**American Indian, Black, Multiracial, White**

### Problem 2 - Evaluating Missing Values

Which variables have at least on interviewee with a missing (NA) value ? **MetroAreaCode, Married, Education, EmploymentStatus, Industry**

We will try to determine if there is a pattern in the missing values of the Married variable. The function
is.na() returns a vector of TRUE/FALSE values for whether the Married variable is missing. We can see if 
Married is missing based on the Region variable

```{r}
table(CPS$Region, is.na(CPS$Married))
table(CPS$Sex, is.na(CPS$Married))
table(CPS$Age, is.na(CPS$Married))
table(CPS$Citizenship, is.na(CPS$Married))
```

**The Married variable being missing is related to the Age value for the interviewee**

MetroAreaCode is missing if an interviewee does not live in a metropolitan area.

```{r}
table(CPS$State, is.na(CPS$MetroAreaCode))
table(CPS$Region, is.na(CPS$MetroAreaCode))
```

How many states had all interviewees living in a non-metropolitan area? **2**

How many states had all interviewees living in a metropolitan area? **3**

Which region of the United States has the largest proportion of interviewees living in a non-metropolitan area?
**Midwest**

There is a less tedious way to compute the proportion of values that are TRUE. The mean() function, which
takes the average of the values passed to it, wille treat TRUE as 1 and FALSE as 0, meaning it returns the
proportion of values that are true. `r mean(c(TRUE, FALSE, TRUE, TRUE)) # 0.75`

```{r}
sort(tapply(is.na(CPS$MetroAreaCode),CPS$State, mean))
```

Which state has a proportion of interviewees living in a non-metropolitan area closest to 30%? **Wisconsin**

Which state has the largest proportion of non-metropolitan interviewees, ignoring states where all 
interviewees were non-metropolitan? **Montana**

### Problem 3 - Integrating Metropolitan Area Data

Read the dictionaries MetroAreaCodes.csv and CountryCodes.csv, which repectively map MetroAreaCode and
CountryOfBirthCode into their true values.

```{r}
MetroAreaMap = read.csv("MetroAreaCodes.csv")
str(MetroAreaMap)
CountryMap = read.csv("CountryCodes.csv")
str(CountryMap)
```

How many obs are there in MetroAreaMap ? **271**

How many obs are there in CountryMap ? **149**

The following command merges the two data frames on these columns, overwriting the CPS data frame with the result

The first two arguments determine the data frames to be merged (x and y). by.x="MetroAreaCode" means
we're matching on the MetroAreaCode variable from the "x" data frame (CPS), while by.y="Code" means
we're matching on the Code variable from the "y" data frame (MetroAreaMap). all.x=TRUE means we want to keep
all rows from the "x" data frame (CPS).

```{r}
CPS = merge(CPS, MetroAreaMap, by.x="MetroAreaCode", by.y="Code", all.x=TRUE)
summary(CPS)
str(CPS)
sort(table(CPS$MetroArea), decreasing = TRUE)
```

What was the name of the variable that was added to the data frame by the merge() operation ? **MetroArea**

How many interviewees have a missing value for the new metropolitan area variable? **34238**

Which of the following metropolitan area has the largest number of interviewees ? **Boston**

```{r}
sort(tapply(CPS$Hispanic, CPS$MetroArea, mean),decreasing = TRUE)
```

Which metropolitan area has the highest proportion of interviewees of Hispanic ethnicity? **Laredo, TX**

```{r}
sort(tapply(CPS$Race == "Asian", CPS$MetroArea, mean),decreasing = TRUE)
```

Number of metropolitan areas in the USA from which at least 20% of interviewees are Asian: **4**

```{r}
sort(tapply(CPS$Education == "No high school diploma", CPS$MetroArea, mean, na.rm = TRUE))
```

Which metropolitan area has the smallest proportion of interviewees who have received no high school diploma ?
**Iowa City, IA**

### Problem 4 - Integrating Country of Birth Data

Merge in the country of birth information from the CountryMap data frame, replacing the CPS data frame with the
result.

```{r}
CPS = merge(CPS, CountryMap, by.x="CountryOfBirthCode", by.y="Code", all.x=TRUE)
summary(CPS)
```

What is the name of the variable added to the CPS data frame by this merge operation? **Country**

How many interviewees have a missing value for the new country of birth variables? **176**

```{r}
sort(table(CPS$Country), decreasing = TRUE)
```

Among all interviewees born outside of the North America, which country was the most common place of birth?
**Philippines**

```{r}
tapply(CPS$Country != "United States", CPS$MetroArea, mean, na.rm = TRUE)
```

What proportion of the interviewees from the "New York-Northern New Jersey-Long Island, NY-NJ-PA" metropolitan
area have a country of birth that is not the US? **0.309**

```{r}
# India
table(CPS$Country == "India", CPS$MetroArea == "Boston-Cambridge-Quincy, MA-NH")
table(CPS$Country == "India", CPS$MetroArea == "Minneapolis-St Paul-Bloomington, MN-WI")
table(CPS$Country == "India", CPS$MetroArea == "New York-Northern New Jersey-Long Island, NY-NJ-PA")
table(CPS$Country == "India", CPS$MetroArea == "Washington-Arlington-Alexandria, DC-VA-MD-WV")
# Brazil
table(CPS$Country == "Brazil", CPS$MetroArea == "Boston-Cambridge-Quincy, MA-NH")
table(CPS$Country == "Brazil", CPS$MetroArea == "Minneapolis-St Paul-Bloomington, MN-WI")
table(CPS$Country == "Brazil", CPS$MetroArea == "New York-Northern New Jersey-Long Island, NY-NJ-PA")
table(CPS$Country == "Brazil", CPS$MetroArea == "Washington-Arlington-Alexandria, DC-VA-MD-WV")
# Somalia
table(CPS$Country == "Somalia", CPS$MetroArea == "Boston-Cambridge-Quincy, MA-NH")
table(CPS$Country == "Somalia", CPS$MetroArea == "Minneapolis-St Paul-Bloomington, MN-WI")
table(CPS$Country == "Somalia", CPS$MetroArea == "New York-Northern New Jersey-Long Island, NY-NJ-PA")
table(CPS$Country == "Somalia", CPS$MetroArea == "Washington-Arlington-Alexandria, DC-VA-MD-WV")

```

Which metropolitan area has the largest number of interviewees with a country birth in India ? **New York**

In Brazil ? **Boston**

In Somalia ? **Minneapolis**


