---
title: "TAE - Unit 1 - Lecture"
author: "GT"
date: "12 avril 2016"
output: html_document
---

## Unit 1 - Lecture

### Getting started with R

R can be used as a basic calculator.

```{r}
8*6
2^16
```

R works in terms of functions and variables.

A function can take in several arguments, or input and returns an output value.

```{r}
sqrt(2)
abs(-65)
```

To save the output of a function, it's possible to saving it to a variable

```{r}
SquareRoot2 = sqrt(2)
SquareRoot2
```

Don't use spaces in variable names, don't start variable names with a number

It's possible to do assignement with '=' or '<-'

```{r}
HoursYear <- 365*24
HoursYear
```

List all variable created in the session

```{r}
ls()
```


### Vectors and Data Frames

A vector is a series of numbers or characters stored as the same object.

You can create a vector in R using the 'c' function (combine).

Create a vector of country names, then enter the life expectancies in the same 
order

```{r}
c(2, 3, 5, 8, 13)
Country = c("Brazil", "China", "India", "Switzerland", "USA")
LifeExpectancy = c(74, 76, 65, 83, 79)
Country
LifeExpectancy
```
To display an element of a vector, use square brackets.

```{r}
Country[1] # To get the first
LifeExpectancy[3] # To get the third
```

Another nice function to create vectors is the 'seq' function.

This give a sequence of numbers from 0 to 100 and the numbers increment of two

```{r}
seq(0, 100, 2)
```

Some algorithms require all of the data to be in a sigle object like a data frame

```{r}
CountryData = data.frame(Country, LifeExpectancy)
CountryData
```

To add another variable to the data

```{r}
CountryData$Population=c(199000, 1390000, 1240000, 7997, 318000)
CountryData
```

To add two new observations

```{r}
Country = c("Australia", "Greece")
LifeExpectancy = c(82, 81)
Population = c(23050, 11125)
```

Then create a new data frame

```{r}
NewCountryData = data.frame(Country, LifeExpectancy, Population)
NewCountryData
```

The rbind() function combines both data frames by stacking the rows.

```{r}
AllCountryData = rbind(CountryData, NewCountryData)
AllCountryData
```
### Loading Data Files

Read the data file WHO.csv and save it in data frame

```{r}
getwd()
WHO = read.csv("WHO.csv")
```

To look at the data, there are two useful commands: str() and summary()

str() show the structure of the data, for each variable, str() gives the name of 
the variable, a description of the type of the variable, followed by a first 
few values of the variable.

summary() for the factor variables it count the number of observations in each 
of the levels or categories. For each of the numerical values, it's calculated the min, max, mean,first quartile, third quartile, median

```{r}
str(WHO)
summary(WHO)
```

It's possible to subset the data with the subset() function. The function take two
arguments, the first is the data frame, the second is the criteria for which
observation we want subset

```{r}
WHO_Europe = subset(WHO, Region == "Europe")
```

To save this new data frame to a csv file, there is the write.csv() function

```{r}
write.csv(WHO_Europe, "WHO_Europe.csv")
```

After the saving, it's possible to remove the data frame from the current session

```{r}
ls()
rm(WHO_Europe)
ls()
```

### Summary Statistics and Scatterplot

To access a variable in a data frame, it has to be linked to the data frame
it belongs to with the dollar sign. The output is a vector of the data frame WHO.
It's possible to compute some statistics about the variable like the mean or
the standard deviation.

```{r}
WHO$Under15
mean(WHO$Under15)
sd(WHO$Under15)
```

It's possible to get the statistical summary of just one variable

```{r}
summary(WHO$Under15)
```

This output tells us that there's a country with only 13% of the population 
under 15.

It's possible to see which country it is using the which.min() function. The
returned number is the row number of the observation and it's easy to see which country is in this row.

```{r}
which.min(WHO$Under15)
WHO$Country[86]
``` 

Example, to see the country that has the maximum percentage of the population 
under 15.

```{r}
WHO$Country[which.max(WHO$Under15)]
```

To create a scatter plot of GNI vs fertility rate using the plot() function

```{r}
plot(WHO$GNI, WHO$FertilityRate)
```

There are few countries for which both the GNI and the fertility rate are high, 
and using the nrow() function to see how many rows are in the subset.

```{r}
Outliers = subset(WHO, GNI > 10000 & FertilityRate > 2.5)
nrow(Outliers)
```

There's an easy way to extract a few variables from a dataset

```{r}
Outliers[c("Country", "GNI", "FertilityRate")]
```

### Plots and Summary Tables

In addition to scatter plots, there are several other types of plots in R.

Create an histogram with the hist() function. A histogram is useful for
understanding the distribution of a variable

```{r}
hist(WHO$CellularSubscribers)
```

It's easy to create a boxplot of Life expecancy sorted by region. A box plot is
useful for understanding the statistical range of a variable. 
The box for each region shows the range between the first and third quartiles with
the middle line marking the median value. 
The dashed lines at the top and bottom of the box show the range from 

- the minimum: first quartile minus 1.5*IQR (inter-quartile range) and
- the maximum: third qurtile plus 1.5*IQR (inter-quartile range) 

excluding the outliers

```{r}
boxplot(WHO$LifeExpectancy ~ WHO$Region)
```

In order to have nice labels it's necessary to add a few arguments

```{r}
boxplot(WHO$LifeExpectancy ~ WHO$Region, ylab = "Life Expectancy", main = "Life Expectancy of Countries by Region")
```

Look at the data using  some summary tables This counts the number of obs in each
category. Table works well for variables with only a few possible values.

```{r}
table(WHO$Region)
```

In order to see some nice information about numerical variables by using tapply()
function. This split the obs by region and then computes the mean of the variable
Over60. Tapply splits the data by the second argument given and then apply the 
third argument function

```{r}
tapply(WHO$Over60, WHO$Region, mean)
# Another example
tapply(WHO$LiteracyRate, WHO$Region, min)
tapply(WHO$LiteracyRate, WHO$Region, min, na.rm=TRUE)
```

The 'na.rm' option removes all missing value before doing the computation