---
title: "Unit 1 - Assignement"
author: "GT"
date: "16 avril 2016"
output: html_document
---

## AN ANALYTICAL DETECTIVE

### Problem 1 - Loading the Data

```{r}
Sys.setlocale("LC_ALL", "C")
mvt = read.csv("mvtWeek1.csv")
str(mvt)
summary(mvt)
```

How many observations ? **191641**

How many variables ? **11**

What is the maximum value of the variable "ID" ? **9181151**

What is the minimum value of the variable "Beat" ? **111**

How many obs have value TRUE in the "Arrest" variables ? **15536**

How many observations have a LocationDescription value of ALLEY ? **2308**

### Problem 2 - Understanding Dates in R

In what format are the entries in the variable Date ? **mm/dd/yy hh:mm**

```{r}
mvt$Date[1]
```

Let's convert these characters into a Date object in R.

```{r}
DateConvert = as.Date(strptime(mvt$Date, "%m/%d/%y %H:%M"))
summary(DateConvert)
```

What's the month of the year of the median date in the dataset ? **May 2006**

Let's extract the month and the day of the week, and add these variables to the data.

```{r}
mvt$Month = months(DateConvert)
mvt$Weekday = weekdays(DateConvert)
mvt$Date = DateConvert
table(mvt$Month)
table(mvt$Weekday)
table(mvt$Month, mvt$Arrest)
```

In which month did the fewest motor vehicle thefts occur ? **February**

On which weekday did the most motor vehicle thefts occur ? **Friday**

Which month has the largest number of motor vehicle thefts for which
an arrest was made ? **January**