---
title: "TAE - Unit 1 - Recitation"
author: "GT"
date: "14 avril 2016"
output: html_document
---

## Unit 1 - Recitation

### The importance of Food and Nutrition

Good nutrition is an important part of leading a healthy lifestyle. Malnutrition can result in obesity.
More than 35% of American adults are obese. The trends Worldwide are no different. Obesity is one of
today's visible public health problems.  
So good nutrition is essential for an overall healthy lifestyle. We have access to hundreds of nutrition and
weight loos applications. These apps are mostly powered by the United States Departement of Agricultural, USDA, 
food database. These database is about over 7000 food items including the amount of calories and nutrients.

### Working with Data in R

Start by reading the dataset `USDA.csv`, which contains all foods in the USDA database in 100 grams amounts.

Use the function read.csv() and save the output to a data frame, USDA
Learn about the data using the functino str()

```{r}
USDA = read.csv("USDA.csv")
str(USDA)
```

The first variable is a unique identification number for each of the foods 
starting with 1001...

To obtain high-level statistical information about the dataset, it can be used the
summary() function

```{r}
summary(USDA)
```

### Data Analysis

The maximum level of Sodium is 38758, which is very high.

To see which food this correspends to, first is necessary to select the vector of 
data associated with the variable Sodium, then find which food is associated with
the function which.max() and use the answer (the index) to look at the vector
corresponding to the text descriptor of the foods

```{r}
head(USDA$Sodium, 20)
which.max(USDA$Sodium)  # 265
```

In order to remember the name of the variable, it can be used the names() function

```{r}
names(USDA)
```

It's possible now to check for the corresponding food

```{r}
USDA$Description[265] # salt, table
```

Find out which foods contain more than 10000 mg of sodium creating a subset 
of the original data frame

```{r}
HighSodium = subset(USDA, Sodium > 10000)
```

Check how many foods exist in this new data frame with the function nrow()

```{r}
nrow(HighSodium) # 10
```

Check the name of foodss by looking at their Description vector.

```{r}
HighSodium$Description
```

Find how much sodium is contained in 100 gr of caviar, get the index of caviar 
first using the function match() and track down the word 'caviar' in the text
description, then find the sodium level.

```{r}
match("CAVIAR", USDA$Description) # 4154
USDA$Sodium[4154] # 1500 mg
```

This sequence of operations can be done with just one expression

```{r}
USDA$Sodium[match("CAVIAR", USDA$Description)]
```

Evaluate this value by comparing it to the mean and the standard deviation of the
sodium levels.

```{r}
summary(USDA$Sodium)
sd(USDA$Sodium)
```

Don't forget to remove the missing values.

```{r}
sd(USDA$Sodium, na.rm = TRUE) # 1045.417
```

If mean and sd are summed `322.1+1045.417 = 1367.517`, 
the result is still smaller than the amount of sodium in 100 gr of caviar

### Creating Plots in R

Visualization is a crucial step for initial data exploration.

First create a scatterplot with Protein on the x-axis and Fat on the y-axis
using the plot() function

```{r}
plot(USDA$Protein,USDA$TotalFat)
```

The plot has a very interesting triangular shape, it looks like foods that are higher in protein are typically lower in fat and vice versa.

R gives default names for the axis. It's possible to modify them adding more
arguments to the plot function, add also a title and change the color

```{r}
plot(USDA$Protein,USDA$TotalFat, xlab="Protein", ylab="Fat", main="Protein vs Fat",col="red")
```

Another way to visualize data is by plotting histograms using the hist() function.

```{r}
hist(USDA$VitaminC, xlab="Vitamin C (mg)", main="Histogram of Vitamin C Levels")
```

The histogram groups them all together in one cell. In order to zoom in this
section, it's necessary to limit the x-axis to go from zero to 100 mg.

```{r}
hist(USDA$VitaminC, xlab="Vitamin C (mg)", main="Histogram of Vitamin C Levels",xlim=c(0,100))
```

But now there is only one big cell. To really break the cell into smaller pieces,
it's necessary to add the argument 'breaks='

```{r}
hist(USDA$VitaminC, xlab="Vitamin C (mg)", main="Histogram of Vitamin C Levels",xlim=c(0,100),breaks=100)
```

There are actually only 5 cells and each cell is 20 mg long. This beacause the
orginal interval, 0 to 2000 is divided into 100 cells, `2000/100=20`.

To correct this it's needed to think in terms of the original interval and break it
into 2000 cells.

```{r}
hist(USDA$VitaminC, xlab="Vitamin C (mg)", main="Histogram of Vitamin C Levels",xlim=c(0,100),breaks=2000)
```

In the refined histogram it's possible to see that  more than 5000 foods have less
than 1 mg of Vitamim C.

A third way to visualize data is using box plots using boxplot() function

```{r}
boxplot(USDA$Sugar, main="Boxplot of Sugar Levels", ylab="Sugar (g)")
```

The average across the data set seems to be pretty low: around 5 gr.
But there are a lot of outliers with extremely high values of sugar
There exist some foods that have almost 100 gr of sugar in 100 gr.


### Adding Variables

Suppose that it's needed to add a variable in the USDA data frame that takes a 
value 1 if the food has higher sodium than average and 0 if the food has lower
sodium than average.

Compare if a food has more sodium than the avearge, for example the first food
Dont't forget to remove the non-available entries.

```{r}
USDA$Sodium[1] > mean(USDA$Sodium, na.rm=TRUE) # TRUE
# What about the 50th food?
USDA$Sodium[50] > mean(USDA$Sodium, na.rm=TRUE) # FALSE
```

It's possible to write the same command on all the vector Sodium and save the results into a vector called HighSodium, then look at its structure.

```{r}
HighSodium = USDA$Sodium > mean(USDA$Sodium, na.rm=TRUE) 
str(HighSodium)
```

The vector has all these values TRUE and FALSE which are called logicals, so the
type of the HighSodium vector is logical.
But it's expected to have 1 and 0 instead of TRUE and FALSE, to do this it exist 
the function as.numeric().

```{r}
HighSodium = as.numeric(USDA$Sodium > mean(USDA$Sodium, na.rm=TRUE))
str(HighSodium)
```

This vector is not yet associated with the USDA data frame. 
To do this it,  use the dollar notation.

```{r}
USDA$HighSodium = as.numeric(USDA$Sodium > mean(USDA$Sodium, na.rm=TRUE))
str(USDA)
```

It's possible to do the same and add the variables HighProtein, HighCarbs, HighFat

```{r}
USDA$HighProtein = as.numeric(USDA$Protein > mean(USDA$Protein, na.rm=TRUE))
USDA$HighFat = as.numeric(USDA$TotalFat > mean(USDA$TotalFat, na.rm=TRUE))
USDA$HighCarbs = as.numeric(USDA$Carbohydrate > mean(USDA$Carbohydrate, na.rm=TRUE))
str(USDA)
```

### Summary tables

Now we can try to understand the data and the relationships between variables using table() and tapply() functions.

To figure out how may foods have higher sodium level than average, look at the HighSodium variable and count the foods that have values 1.

```{r}
table(USDA$HighSodium) # most of the food have lower sodium than average
```

Look at how many foods have both high sodium and high fat

```{r}
table(USDA$HighSodium, USDA$HighFat)
```

The rows belong to the first input (HighSodium) and the columns correspond to the second input (HighFat).
Compute the average amount of iron sorted by high and low protein.

The tapply function takes three arguments:

`tapply(argument1, argument2, argument3)`

Group argument1 by argument2 and apply argument3

```{r}
tapply(USDA$Iron, USDA$HighProtein, mean, na.rm=TRUE)
```

How about the maximum level of vitamin C in food with high and low carbs.

```{r}
tapply(USDA$VitaminC, USDA$HighCarbs, max, na.rm=TRUE)
```

In order to further analyze these data use the summary function

```{r}
tapply(USDA$VitaminC, USDA$HighCarbs, summary, na.rm=TRUE)
```