---
title: "Unit 1 - Assignement"
author: "GT"
date: "16 avril 2016"
output: html_document
---

## AN ANALYTICAL DETECTIVE

### Problem 1 - Loading the Data

```{r}
Sys.setlocale("LC_ALL", "C")
mvt = read.csv("mvtWeek1.csv")
str(mvt)
summary(mvt)
```

How many observations ? **191641**

How many variables ? **11**

What is the maximum value of the variable "ID" ? **9181151**

What is the minimum value of the variable "Beat" ? **111**

How many obs have value TRUE in the "Arrest" variables ? **15536**

How many observations have a LocationDescription value of ALLEY ? **2308**

### Problem 2 - Understanding Dates in R

In what format are the entries in the variable Date ? **mm/dd/yy hh:mm**

```{r}
mvt$Date[1]
```

Let's convert these characters into a Date object in R.

```{r}
DateConvert = as.Date(strptime(mvt$Date, "%m/%d/%y %H:%M"))
summary(DateConvert)
```

What's the month of the year of the median date in the dataset ? **May 2006**

Let's extract the month and the day of the week, and add these variables to the data.

```{r}
mvt$Month = months(DateConvert)
mvt$Weekday = weekdays(DateConvert)
mvt$Date = DateConvert
table(mvt$Month)
table(mvt$Weekday)
table(mvt$Month, mvt$Arrest)
```

In which month did the fewest motor vehicle thefts occur ? **February**

On which weekday did the most motor vehicle thefts occur ? **Friday**

Which month has the largest number of motor vehicle thefts for which
an arrest was made ? **January**

### Problem 3 - Visualizing Crime Trends

Make a histogram of the variable Date. Add an extra argument, to specify the number of bars we want into histogram

```{r}
hist(mvt$Date, breaks=100)
```

Does it look like crime increase or decrease from 2002-2012 ? **Decrease**

Does it look like crime increase or decrease frome 2005-2008 ? **Decrease**

Does it look like crime increase or decrease from 2009-2011 ? **Increase**

Create a boxplot of the variable "Date", sorted by the variable "Arrest"

```{r}
boxplot(mvt$Date ~ mvt$Arrest)
```

Does it look like there were more crimes for which arrests were made in the first half of the time period or 
the second half ? **First half**

```{r}
table(mvt$Year, mvt$Arrest)
Arrest2001 = subset(mvt, mvt$Year == 2001 & mvt$Arrest == TRUE)
Total2001 = subset(mvt, mvt$Year == 2001)
nrow(Arrest2001)/nrow(Total2001)
Arrest2007 = subset(mvt, mvt$Year == 2007 & mvt$Arrest == TRUE)
Total2007 = subset(mvt, mvt$Year == 2007)
nrow(Arrest2007)/nrow(Total2007)
Arrest2012 = subset(mvt, mvt$Year == 2012 & mvt$Arrest == TRUE)
Total2012 = subset(mvt, mvt$Year == 2012)
nrow(Arrest2012)/nrow(Total2012)
```

For what proportion of motor vehicle thefts in 2001 was an arrest made? **0.104**

For what proportion of motor vehicle thefts in 2007 was an arrest made? **0.085**

For what proportion of motor vehicle thefts in 2012 was an arrest made? **0.039**

### Problem 4 - Popular Locations

Find the top five locations where motor vehicle thefts occur more by using the sort function:
**Street, Parking Lot/Garage (Non. Resid.), Alley, Gas Station, and Driveway - Residential**

```{r}
sort(table(mvt$LocationDescription),decreasing = TRUE)
```

Create a subset of your data, only taking observations for which the theft happened in one of these five
locations

```{r}
Top5 = subset(mvt, mvt$LocationDescription %in% c("STREET", "PARKING LOT/GARAGE(NON.RESID.)", "ALLEY", "GAS STATION", "DRIVEWAY - RESIDENTIAL"))
nrow(mvt)
nrow(Top5)
```

How many obs are in Top5 **177510**

R will remember the other categories of the LocationDescription variable from the orginal dataset. 
To make the tables nicer we can refresh this factor variable

```{r}
Top5$LocationDescription = factor(Top5$LocationDescription)
table(Top5$LocationDescription, Top5$Arrest)
table(Top5$Weekday, Top5$LocationDescription)
```
One of the locations has a much higher arrest rate than the other locations. Which is it ? **Gas Station**

On which day of the week do the most motor vehicle thefts at gas stations happen ? **Saturday**

On which day of the week do the fewest motor vehicle thefts in residential driveway happen? **Saturday**
