---
title: "Unit 3 - Assignment 1"
author: "GT"
date: "May 2, 2016"
output: html_document
---

## Popularity of music records

How can we use analytics to predict the popularity of a song? In this assignment, we challenge ourselves
to predict whether a song will reach a spot in the Top 10 of the Billboard Hot 100 Chart.

Taking an analytics approach, we aim to use information about a song's properties to predict its
popularity. The dataset songs.csv consists of all songs which made it to the Top 10 of the Billboard Hot
100 Chart from 1990-2010 plus a sample of additional songs that didn't make the Top 10.

The variables included in the dataset either describe the artist or the song, or they are associated 
with the following song attributes: time signature, loudness, key, pitch, tempo, and timbre.

Here's a detailed description of the variables:

- *year*: the year the song was released
- *songtitle*: the title of the song
- *artistname*: the name of the artist of the song
- *songID* and *artistID*: identifying variables for the song and artist
- *timesignature* and *timesignature_confidence*: a variable estimating the time signature of the song,
  and the confidence in the estimate
- *loudness*: a continuous variable indicating the average amplitude of the audio in decibels
- *tempo* and *tempo_confidence*: a variable indicating the estimated beats per minute of the song, 
  and the confidence in the estimate
- *key* and *key_confidence*: a variable with twelve levels indicating the estimated key of the song 
  (C, C#, . . ., B), and the confidence in the estimate
- *energy*: a variable that represents the overall acoustic energy of the song, using a mix of features
  such as loudness
- *pitch*: a continuous variable that indicates the pitch of the song
- *timbre_0_min*, *timbre_0_max*, *timbre_1_min*, *timbre_1_max*, . . . , *timbre_11_min*, and
  *timbre_11_max*: variables that indicate the minimum/maximum values over all segments for each of the
  twelve values in the timbre vector (resulting in 24 continuous variables)
- *Top10*: a binary variable indicating whether or not the song made it to the Top 10 of the Billboard
  Hot 100 Chart (1 if it was in the top 10, and 0 if it was not)

### Problem 1

```{r}
songs = read.csv("songs.csv")
str(songs)
table(songs$year)
MichaelJackson = subset(songs,songs$artistname == "Michael Jackson")
nrow(MichaelJackson)
subset(MichaelJackson, MichaelJackson$Top10 == 1)[c("songtitle", "Top10")]
table(songs$timesignature)
songs$songtitle[which.max(songs$tempo)]
```

### Problem 2

```{r}
SongsTrain = subset(songs, songs$year < 2010)
SongsTest = subset(songs, songs$year == 2010)
str(SongsTrain)
str(SongsTest)
nonvars = c("year", "songtitle", "artistname", "songID", "artistID")
SongsTrain = SongsTrain[, !(names(SongsTrain) %in% nonvars)]
SongsTest = SongsTest[, !(names(SongsTest) %in% nonvars)]
SongsLog1 = glm(Top10 ~ ., data = SongsTrain, family=binomial)
summary(SongsLog1)
```

### Problem 3

```{r}
cor(SongsTrain$loudness, SongsTrain$energy)
SongsLog2 = glm(Top10 ~ . -loudness, data = SongsTrain, family=binomial)
summary(SongsLog2)
SongsLog3 = glm(Top10 ~ . -energy, data = SongsTrain, family=binomial)
summary(SongsLog3)
```

### Problem 4

```{r}
predTest = predict(SongsLog3, type = "response", newdata = SongsTest)
table(SongsTest$Top10, predTest > 0.45)
(309+19)/(309+19+5+40)
(309+5)/(309+19+5+40) # baseline
# sensitivity
19/(19+40)
# specificity
309/(309+5)
```

