---
title: "Unit 3 - Assignment 3"
author: "GT"
date: "2 mai 2016"
output: html_document
---

## Predicting loan repayment

This dataset represents 9,578 3-year loans that were funded through the LendingClub.com platform between 
May 2007 and February 2010. The binary dependent variable **not_fully_paid** indicates that the loan was 
not paid back in full (the borrower either defaulted or the loan was "charged off," meaning the borrower 
was deemed unlikely to ever pay it back).

To predict this dependent variable, we will use the following independent variables available to the investor when deciding whether to fund a loan:

- *credit.policy*: 1 if the customer meets the credit underwriting criteria of LendingClub.com, 
   and 0 otherwise.
- *purpose*: The purpose of the loan (takes values "credit_card", "debt_consolidation", 
  "educational", "major_ purchase","small_business", and "all _other").
- *int.rate*: The interest rate of the loan, as a proportion (a rate of 11% would be stored as 0.11). 
  Borrowers judged by LendingClub.com to be more risky are assigned higher interest rates.
- *installment*: The monthly installments ($) owed by the borrower if the loan is funded.
- *log.annual.inc*: The natural log of the self-reported annual income of the borrower.
- *dti*: The debt-to-income ratio of the borrower (amount of debt divided by annual income).
- *fico*: The FICO credit score of the borrower.
- *days.with.cr.line*: The number of days the borrower has had a credit line.
- *revol.bal*: The borrower's revolving balance (amount unpaid at the end of the credit card billing cycle).
- *revol.util*: The borrower's revolving line utilization rate (the amount of the credit line used 
  relative to total credit available).
- *inq.last.6mths*: The borrower's number of inquiries by creditors in the last 6 months.
- *delinq.2yrs*: The number of times the borrower had been 30+ days past due on a payment in the past 2 years.
- *pub.rec*: The borrower's number of derogatory public records (bankruptcy filings, tax liens, or judgments).

### Problem 1

```{r}
loans = read.csv("loans.csv")
str(loans)
summary(loans)
table(loans$not.fully.paid)
1533/(1533+8045)
loans = read.csv("loans_imputed.csv")
```

### Problem 2

```{r}
set.seed(144)
library(caTools)
split = sample.split(loans$not.fully.paid, SplitRatio = 0.7)
train = subset(loans, split == TRUE)
test = subset(loans, split == FALSE)
Mod1 = glm(not.fully.paid ~ ., data = train, family = binomial)
summary(Mod1)
A = (-9.317e-3)*700
A
B = (-9.317e-3)*710
B
A-B
exp(A)/exp(B)
predicted.risk = predict(Mod1, newdata = test, type = "response")
summary(predicted.risk)
test$predicted.risk = predicted.risk
table(test$not.fully.paid, predicted.risk > 0.5)
# accuracy
(2400+3)/(2400+3+457+13)
# baseline
(2400+13)/(2400+3+457+13)
library(ROCR)
ROCRpred = prediction(predicted.risk, test$not.fully.paid)
ROCRperf = as.numeric(performance(ROCRpred, "auc")@y.values)
ROCRperf
```

### Problem 3

```{r}
Mod2 = glm(not.fully.paid ~ int.rate, data = train, family = binomial)
summary(Mod2)
cor(train[c("int.rate","installment","log.annual.inc","dti","fico","days.with.cr.line","revol.bal","revol.util","credit.policy","inq.last.6mths","delinq.2yrs","pub.rec")])
predicted.risk = predict(Mod2, newdata = test, type = "response")
summary(predicted.risk)
table(test$not.fully.paid, predicted.risk > 0.5)
ROCRpred = prediction(predicted.risk, test$not.fully.paid)
ROCRperf = as.numeric(performance(ROCRpred, "auc")@y.values)
ROCRperf
```

